--- 
title: "Biostatistics using R"
author: "Dewey Brooke"
date: "`r Sys.Date()`"
knit: "bookdown::render_book"
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
colorlinks: yes
lot: yes
lof: yes
fontsize: 12pt
monofont: "Source Code Pro"
monofontoptions: "Scale=0.7"
github-repo: dbrooke/bst4r
always_allow_html: yes
site: bookdown::bookdown_site
description: ""
---

```{r setup, include=FALSE}
options(
  htmltools.dir.version = FALSE, formatR.indent = 2,
  width = 55, digits = 4, warnPartialMatchAttr = FALSE, warnPartialMatchDollar = FALSE
)

knitr::opts_chunk$set(comment = NULL, cache = TRUE)

```

# Preface {-}

This is a _sample_ book written in **Markdown**. You can use anything that Pandoc's Markdown supports, e.g., a math equation $a^2 + b^2 = c^2$.

The **bookdown** package can be installed from CRAN or Github:

```{r eval=FALSE}
install.packages("bookdown")
# or the development version
# devtools::install_github("rstudio/bookdown")
```

Remember each Rmd file contains one and only one chapter, and a chapter is defined by the first-level heading `#`.

To compile this example to PDF, you need XeLaTeX. You are recommended to install TinyTeX (which includes XeLaTeX): <https://yihui.name/tinytex/>.

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
